% LTeX: enabled=false

\documentclass[sigplan,11pt,screen,nonacm]{acmart}
\settopmatter{printfolios}

\usepackage{lipsum}
\usepackage{amsmath}

\usepackage{subcaption}
\usepackage{hyphenat}
\usepackage[capitalise, nameinlink, noabbrev]{cleveref}
\usepackage{minted}
\setminted{autogobble, numbersep=0.5em, escapeinside=||}
\usemintedstyle{friendly}
\usepackage[babel]{csquotes}

\usepackage{tikz}
\usetikzlibrary{tikzmark, calc, arrows.meta}

\usepackage{bussproofs}
\usepackage{simplebnf}

\newcommand*{\Copy}{\mintinline{text}{Copy}}

\newcounter{codeline}
\newcommand*{\codelinelabel}[1]{
  \refstepcounter{codeline}
  \setcounter{codeline}{\value{FancyVerbLine}}
  \label[codeline]{#1}
}
\crefname{codeline}{line}{lines}
\Crefname{codeline}{Line}{Lines}

\begin{document}
\title{Linear Types and a Lifetime of Borrows}
% LTeX: dictionary+=Lindae
\author{Thomas Lindae}
\affiliation{
  \institution{Technical University of Munich}
  \country{Germany}
}
\email{thomas.lindae@in.tum.de}

% LTeX: enabled=true

\begin{abstract}
  Rust’s type system provides strong memory safety guarantees while enabling low-level systems programming through the notions of ownership, borrowing and lifetimes, which employ strict rules on what programs are valid. These rules are checked by Rust's borrow checker, and draw on the theoretical foundations of linear types. This paper presents a high-level overview of the theoretical foundations of linear types, as well as their application in Rust’s type system. Additionally, we examine practical considerations and current limitations of the borrow checker. We begin by introducing intuitionistic logic and illustrate its connection to type systems through the Curry-Howard correspondence. We then repeat this process for linear logic and linear types. Furthermore, we examine Rust as a real-world application of these concepts, highlighting ownership, capability-passing style, borrowing, and lifetimes, and explore modern developments in Rust’s compiler, including non-lexical lifetimes and the Polonius borrow checker. By bridging theory and practice, this work highlights the principles underlying Rust’s approach to safe and efficient systems programming.
\end{abstract}

\maketitle

\sloppy

\section{Introduction} \label{sec:introduction}

\(\sim\)65000 lines of Rust code have already made it into the Linux kernel source code~\cite{intro:rust-in-linux} all the while Ubuntu is considering replacing the GNU coreutils with uutils, a Rust rewrite~\cite{intro:uutils, intro:ubuntu}. It is clear that the Rust programming language continues to gain in relevance. These accomplishments are, in part, due to Rust's unique blend of low-level systems programming, safety guarantees and an expressive type system.

The latter two are somewhat connected, as Rust's type system is a major part of how Rust can manage its memory safety guarantees. Unless Rust's \mintinline{text}{unsafe} keyword is used, the Rust compiler guarantees to deny programs that would cause certain typical memory safety vulnerabilities, including data races and use-after-free.

At the heart of Rust's type system lie the concepts of \emph{ownership}, \emph{borrowing} and \emph{lifetimes}, whose rules are checked by a component of the compiler called the \emph{borrow checker}. These concepts have since been adopted in various forms in other languages, like OCaml~\cite{intro:ocaml-ownership} and Swift~\cite{intro:swift-ownership}.

Theoretical formalizations of Rust have been developed by \citet{rust:oxide}, \citet{rust:rustbelt}, while \citet{from-linearity-to-borrowing} analyse these concepts in a more general light. But the core mathematical background behind ownership is that of linear logic, originally developed by \citet{girard-linear-logic}.

This paper gives a more high level view of the entire line from the theoretical type-theoretic background, over Rust's basic implementation of these theoretical concepts, all the way to cutting-edge modern developments on the Rust compiler.

To that end, we will first introduce the theoretical background of intuitionistic logic and its correspondence to type systems through the Curry-Howard isomorphism (\cref{sec:intuitionistic:logic} and \cref{sec:intuitionistic:types}). We will further examine linear logic and apply a similar isomorphism to introduce the notion of linear types (\cref{sec:linear:logic} and \cref{sec:linear:types}), with which we can impose some limitations on what programs are valid in exchange for safer and clearer resource management.

We will then take Rust's ownership model as a real world application of linear types (\cref{sec:rust:ownership}). Through motivational examples, we will introduce borrowing, a mechanism that allows escaping the limitations of linearity in exchange for limiting certain mutations (\cref{sec:rust:capability-passing-style} and \cref{sec:rust:borrowing}).

In order to retain safety with borrowing, we will find that the notion of \emph{lifetimes} is required (\cref{sec:rust:lifetimes}). Rust's lifetimes are a complex topic with a lot of history of improvements. Each iteration of how lifetimes are processed in Rust attempts to widen the set of programs accepted by the compiler, and ultimately reduce the friction and complications programmers face when writing Rust code, without compromising Rust's memory safety guarantees.

The two most prominent developments in the space of Rust lifetimes have been \emph{non-lexical-lifetimes}, discussed in \cref{sec:rust:nll}, and the more recent Polonius, discussed in \cref{sec:rust:polonius}.

\section{The Theory of Logic and Types} \label{sec:theory}

Type systems and mathematical logic have unintuitively deep ties. For example, the notion of type inference in modern languages is based mostly on Hindley-Milner type systems, its name originating from the fact that Hindley (a logician) and Milner (a computer scientist) found the same system independently of one another~\cite{hindley, milner}.

While we will not look at type inference in this work, and will instead focus on type systems in general, it gives an early intuition that these topics are interlinked.

\subsection{Intuitionistic Logic} \label{sec:intuitionistic:logic}

Logic provides formal languages and inference rules for precise reasoning. When logic is formalized, proofs become mathematical objects that can be analyzed. Intuitionistic logic is one such formal system. It's primary differentiator from classical logic is the lack of the law of excluded middle. In classical logic, for any proposition \(A\), the formula \(A \lor \lnot A\) is a tautology. In intuitionistic logic, this is not generally provable.~\cite{mints-intro-intuitionistic-logic}

With this limitation, intuitionistic logic is a logic of constructive proofs. Classical logic allows proofs of a proposition that don't actually construct a specific witness, proofs by contradiction being the typical culprit. In intuitionistic logic, one must explicitly provide the corresponding constructive evidence.

Since we are mostly concerned with general concepts, we will only view a reduced version of the logic used by \citet{wadler-linear-taste}. Specifically, we exclude \emph{disjunctive} connectives (\(\lor\)):
\begin{center}
  \begin{bnf}
    \(A, B, C\) : Proposition ::= \(X\) : Variable | \(A \rightarrow B\) : Implication | \(A \land B\) : Conjunction
  \end{bnf}
\end{center}

\begin{figure*}
  \begin{prooftree}
    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(A \vdash A\)}
    \DisplayProof\qquad
    \AxiomC{\(\Delta, \Gamma \vdash A\)}
    \RightLabel{Exchange}
    \UnaryInfC{\(\Gamma, \Delta \vdash A\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma, A, A \vdash B\)}
    \RightLabel{Contraction}
    \UnaryInfC{\(\Gamma, A \vdash B\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma \vdash B\)}
    \RightLabel{Weakening}
    \UnaryInfC{\(\Gamma, A \vdash B\)}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{\(\Gamma, A \vdash B\)}
    \RightLabel{\(\rightarrow\)-I}
    \UnaryInfC{\(\Gamma \vdash A \rightarrow B\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma \vdash A \rightarrow B\)}
    \AxiomC{\(\Delta \vdash A\)}
    \RightLabel{\(\rightarrow\)-E}
    \BinaryInfC{\(\Gamma, \Delta \vdash B\)}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{\(\Gamma \vdash A\)}
    \AxiomC{\(\Delta \vdash B\)}
    \RightLabel{\(\land\)-I}
    \BinaryInfC{\(\Gamma, \Delta \vdash A \land B\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma \vdash A \land B\)}
    \AxiomC{\(\Delta, A, B \vdash C\)}
    \RightLabel{\(\land\)-E}
    \BinaryInfC{\(\Gamma, \Delta \vdash C\)}
  \end{prooftree}
  \caption{Intuitionistic Logic -- Natural Deduction} \label{fig:intuitionistic:logic:nd}
\end{figure*}

We write \(\Gamma \vdash A\) (read \enquote{\(\Gamma\) entails \(A\)}), in which \(\Gamma\) is some set of propositions (which we will call context) and \(A\) is a single proposition, to convey that, given that the propositions in \(\Gamma\) are true, \(A\) is also true. To do logical reasoning, we need rules for such reasoning, for which we introduce a system that is loosely based on Natural Deduction, seen in \cref{fig:intuitionistic:logic:nd}.

We have the \emph{axiom} Id, the three \emph{structural} rules Exchange, Contraction and Weakening, and \emph{logical} rules for \emph{implications} (\(\rightarrow\)-I and \(\rightarrow\)-E) and \emph{conjunctives} (\(\land\)-I and \(\land\)-E). Note that explicit structurual rules are unusual for Natural Deduction, as they are typically absorbed into the logical rules, but explicit ones will make the transition to linear types simpler.

A rule describes a step of logical reasoning. The core idea is that, when the things above the line hold, the things below the line hold as well. Hence, there are two ways to interpret a rule:
\begin{description}
  \item[Top to bottom]
    If I know the thing at the top holds, I can deduce that the thing at the bottom holds
  \item[Bottom to top]
    If I want to show the thing on the bottom, I first need to show the thing at the top
\end{description}
The second interpretation becomes particularly relevant when writing proofs. One would start with the statement one wants to prove, and by applying the rules of natural deduction from bottom to top, it becomes clear what needs to be proved first. The first interpretation on the other hand is how we normally convey proofs to others. One starts with the axioms or known facts, and keeps deriving new facts until the goal is reached.

Actual proofs are usually written as proof trees, where each branch is an instantiation of a rule and leaves are axioms. For example, \cref{fig:intuitionistic:logic:proof} shows the tree that proves \(A \rightarrow B, A \vdash A \land B\).

\begin{figure}[H]
  \begin{prooftree}
    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(A \vdash A\)}

    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(A \rightarrow B \vdash A \rightarrow B\)}

    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(A \vdash A\)}

    \RightLabel{\(\rightarrow\)-E}
    \BinaryInfC{\(A \rightarrow B, A \vdash B\)}

    \RightLabel{\(\land\)-I}
    \BinaryInfC{\(A, A \rightarrow B, A \vdash A \land B\)}

    \RightLabel{Exchange}
    \UnaryInfC{\(A \rightarrow B, A, A \vdash A \land B\)}

    \RightLabel{Contraction}
    \UnaryInfC{\(A \rightarrow B, A \vdash A \land B\)}
  \end{prooftree}
  \caption{Proof of \(A \rightarrow B, A \vdash A \land B\)} \label{fig:intuitionistic:logic:proof}
\end{figure}

Of particular interest for us are the Contraction and Weakening rules. Contraction being the ability to duplicate truths as much as we like, while Weakening is the ability to forget truths.

\subsection{Intuitionistic Types} \label{sec:intuitionistic:types}

\begin{figure*}
  \begin{subfigure}{0.58\textwidth}
    \centering
    \begin{bnf}
      \(s,t,u,v,w\) ::= \(x\) : Variable
      | \(\lambda x\ldotp s\) : Function Definition | \(s\ t\) : Function Application
      | \((u, v)\) : Tuple Definition | let \((x, y) = t\); \(w\) : Tuple Destructuring
    \end{bnf}
    \caption{Intuitionistic Terms}
  \end{subfigure}
  \vrule
  \begin{subfigure}{0.38\textwidth}
    \centering
    \begin{bnf}
      \(\tau, \pi, \omega\) ::= \(T\) : Base Type
      | \(\tau \rightarrow \pi\) : Functions
      | \(\pi \times \omega\) : Tuple
    \end{bnf}
    \caption{Intuitionistic Types}
  \end{subfigure}
  \caption{Grammar of Typed Lambda Calculus} \label{fig:intuitionistic:types:grammar}
\end{figure*}

We will now move away from logic and into the theory of computation and types. To talk about types we first need a programming language, for which we will use a small, limited and typed lambda calculus, the grammar of which can be seen in \cref{fig:intuitionistic:types:grammar}. We write \(t : \alpha\) to denote that term \(t\) has type \(\alpha\).

Similarly to when we defined natural deduction rules to define the notion of logical reasoning, here we will define typing judgement rules which outline what types certain terms have. We write \(\Gamma \vdash t : \tau\) to say that, under the typing context \(\Gamma\), the term \(t\) has type \(\tau\) (and hence \(t\) is well typed). Typing contexts are a list of assignments from variables to types \(x_0 : \tau_0, x_1 : \tau_1 \dots x_n : \tau_n\). The judgement rules can be seen in \cref{fig:intuitionistic:types:judgements}.

\begin{figure*}
  \begin{prooftree}
    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(x : \tau \vdash x : \tau\)}
    \DisplayProof\qquad
    \AxiomC{\(\Delta, \Gamma \vdash t : \tau\)}
    \RightLabel{Exchange}
    \UnaryInfC{\(\Gamma, \Delta \vdash t: \tau\)}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{\(\Gamma, y : \tau, z : \tau \vdash u : \pi\)}
    \RightLabel{Contraction}
    \UnaryInfC{\(\Gamma, x : \tau \vdash u[x/y,x/z] : \pi\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma \vdash s : \pi\)}
    \RightLabel{Weakening}
    \UnaryInfC{\(\Gamma, x : \tau \vdash s : \pi\)}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{\(\Gamma, x : \tau \vdash t : \pi\)}
    \RightLabel{\(\rightarrow\)-I}
    \UnaryInfC{\(\Gamma \vdash \lambda x\ldotp t : \tau \rightarrow \pi\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma \vdash s : \tau \rightarrow \pi\)}
    \AxiomC{\(\Delta \vdash t : \tau\)}
    \RightLabel{\(\rightarrow\)-E}
    \BinaryInfC{\(\Gamma, \Delta \vdash s\ t : \pi\)}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{\(\Gamma \vdash u : \tau\)}
    \AxiomC{\(\Delta \vdash v : \pi\)}
    \RightLabel{\(\times\)-I}
    \BinaryInfC{\(\Gamma, \Delta \vdash (u, v) : \tau \times \pi\)}
    \DisplayProof\qquad
    \AxiomC{\(\Gamma \vdash t : \tau \times \pi\)}
    \AxiomC{\(\Delta, x : \tau, y : \pi \vdash w : \omega\)}
    \RightLabel{\(\times\)-E}
    \BinaryInfC{\(\Gamma, \Delta \vdash \text{ let } (x, y) = t; w : \omega\)}
  \end{prooftree}
  {\small The notation \(u[x/y,x/z]\) means to replace all occurrences of \(y\) and \(z\) in \(u\) by \(x\).}
  \caption{Intuitionistic Types – Judgement Rules} \label{fig:intuitionistic:types:judgements}
\end{figure*}

As before, we can use these judgements to type check terms (under typing contexts) like \[f : \tau \rightarrow \pi, x : \tau \vdash (x, f\ x) : \tau \times \pi\] shown in \cref{fig:intuitionistic:types:proof}.

\begin{figure*}
  \begin{prooftree}
    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(z : \tau \vdash z : \tau\)}

    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(f : \tau \rightarrow \pi \vdash f : \tau \rightarrow \pi\)}

    \AxiomC{}
    \RightLabel{Id}
    \UnaryInfC{\(y : \tau \vdash y : \tau\)}

    \RightLabel{\(\rightarrow\)-E}
    \BinaryInfC{\(f : \tau \rightarrow \pi, z : \tau \vdash f\ z : \pi\)}

    \RightLabel{\(\times\)-I}
    \BinaryInfC{\(y : \tau, f : \tau \rightarrow \pi, z : \tau \vdash (y, f\ z) : \tau \times \pi\)}

    \RightLabel{Exchange}
    \UnaryInfC{\(f : \tau \rightarrow \pi, y : \tau, z : \tau \vdash (y, f\ z) : \tau \times \pi\)}

    \RightLabel{Contraction}
    \UnaryInfC{\(f : \tau \rightarrow \pi, x : \tau \vdash (x, f\ x) : \tau \times \pi\)}
  \end{prooftree}
  \caption{Proof of \(f : \tau \rightarrow \pi, x : \tau \vdash (x, f\ x) : \tau \times \pi\)} \label{fig:intuitionistic:types:proof}
\end{figure*}

\subsubsection{Curry-Howard correspondence} \label{sec:intuitionistic:curry-howard}

One may notice that the judgement rules in \cref{fig:intuitionistic:types:judgements} and the proof rules in \cref{fig:intuitionistic:logic:nd} are effectively identical. Similarly, the example judgement presented in \cref{fig:intuitionistic:types:proof} and the example proof in \cref{fig:intuitionistic:logic:proof} are also identical. This is the discovery made by Curry and Howard~\cite{curry-combinatory-logic, howard-formalae}: propositions are types, and terms are proofs. The conjunction (\(A \land B\)) for propositions are pairs (\(\tau \times \pi\)) in the world of terms and types, while implications (\(A \rightarrow B\)) are functions (\(\tau \rightarrow \pi\)).

One remarkable point is that terms uniquely represent a proof tree (modulo certain commutative properties~\cite{wadler-linear-taste}). This correspondence isn't quite as obvious as the others, however, the example we used so far can give an intuition on why this is true: The term \((x, f\ x)\) gives significant insight into what a proof of \[A \rightarrow B, A \vdash A \land B\] needs to do. The fact that the variable \(x\) appears twice in the term means there must be a use of the \emph{Contraction} rule, the fact the term is a tuple definition means there must be a use of the \(\land\)\nobreakdash-I rule, and the existence of the function application \(f\ x\) means the proof contains a use of the \(\rightarrow\)\nobreakdash-E rule. The full connection is a bit more involved, so we will leave it at this intuition.

We mentioned how intuitionistic logic is a system of constructive proofs in \cref{sec:intuitionistic:logic}. With the Curry-Howard correspondence, this idea becomes even more concrete: In order to prove a proposition, we must be able to construct a term of the corresponding type. Such a term is a program that, when executed, constructs a value of this type.

\subsection{Linear Logic} \label{sec:linear:logic}

We will shortly step back from type systems into logic again to introduce linear logic, before using the same Curry-Howard isomorphism from \cref{sec:intuitionistic:curry-howard}, to translate linear logic into linear types.

Linear logic is the idea that not all truths are free. With intuitionistic logic, we can duplicate and forget truths as much as we want. In a mathematical sense, if one proves the \emph{Maximum Modulus Theorem}, that theorem can be further used to proof \emph{Schwarz's Lemma} as well as the \emph{Phragmén-Lindelöf Theorem}~\cite{conway-complex}. We don't lose access to the \emph{Maximum Modulus Theorem} after proving \emph{Schwarz's Lemma}.

However, duplication is not free in the real world. If \(A\) means \enquote{I have 60 rupees}, \(B\) means \enquote{I can buy 30 arrows} (that cost 60 rupees total) and \(C\) means \enquote{I can buy a Deku Shield} (which costs 40 rupees), then the following shows a proof of how 60 rupees can buy both in intuitionistic logic by duplicating \(A\):
\begin{center}
  \begin{prooftree}
    \AxiomC{\(A \vdash B\)}
    \AxiomC{\(A \vdash C\)}
    \RightLabel{\(\land\)-I}
    \BinaryInfC{\(A, A \vdash B \land C\)}
    \RightLabel{Contraction}
    \UnaryInfC{\(A \vdash B \land C\)}
  \end{prooftree}
\end{center}
And yet, 60 rupees aren't actually enough to buy both 30 arrows \emph{and} a Deku Shield.

The issue is with the Contraction and Weakening rules, both of which linear logic tries to limit. One could just remove Contraction and Weakening outright, however that would lose us expressivity. Unfortunately, this is where the details get a bit messy, thus we will only give an overview.

The core idea is to split assumptions into two groups: linear and intuitionistic. Intuitionistic assumptions, denoted just as before, work as we know and allow for Contraction and Weakening, while linear assumptions, denoted \(\langle A\rangle\), do not allow Contraction and Weakening.

The notation and meaning of the logical connectives also changes. \(A \rightarrow B\) becomes \(A \multimap B\) (called \enquote{lolli}) and whereas before we could freely convert \(A\) into \(B\), the lolli instead consumes the \(A\). \(A \land B\) becomes \(A \otimes B\) (called \enquote{tensor}) and requires that both sub-propositions be used, meaning \(A \otimes B \vdash A\) is not provable.

Similarly, for our example proof from \cref{fig:intuitionistic:logic:proof}, if we make the assumption \(A\) linear, we get: \[A \multimap B, \langle A\rangle \vdash A \otimes B\] But this isn't provable any more, as we would need to duplicate \(\langle A\rangle\).

\citet{wadler-linear-taste} also explores a second alternative definition of conjunction, which happens to be equivalent to our definition in intuitionistic logic, but gets a profoundly different meaning once it is translated into linear logic.
% TODO maybe write a footnote with more detail, or do an appendix with a full explanation?

\subsection{Linear Types} \label{sec:linear:types}

Applying the Curry-Howard correspondence to linear logic, we get a particularly useful type-theoretic concept: Non-duplicatable and non-forgettable values. Looking at the judgement from \cref{fig:intuitionistic:types:proof} once again: \[f : \tau \rightarrow \pi, x : \tau \vdash (x, f\ x) : \tau \times \pi\] The thing of interest is the duplication of \(x\). In real world programming, \(x\) could be anything, including a highly limited resource. If \(x\) is a simple integer, it may be no problem to copy it, but if \(x\) is a large array, copying may be an unacceptable performance hit. Even worse, if \(x\) represents some resource like a file, duplication may be catastrophic.

The particularly interesting application is in contexts of manual memory management. If we define the following functions inspired by \citet{from-linearity-to-borrowing} for allocating and freeing memory, we may be susceptible to various memory related bugs:

\begin{align*}
  &\vdash \texttt{alloc} : \tau \rightarrow *\tau &
  &\vdash \texttt{load} : *\tau \rightarrow \tau \\
  &\vdash \texttt{free} : *\tau \rightarrow () &
  &\vdash \texttt{store} : *\tau \rightarrow \tau \rightarrow ()
\end{align*}
\(*\tau\) represents a pointer to \(\tau\), and \(()\) represents the unit type. Interpreting these functions in an intuitionistic manner, we can easily produce a use after free bug:\footnote{\mintinline{text}{let y = t; s} is short for \((\lambda y\ldotp s)\ t\)}
\begin{minted}{text}
  let x = alloc v;
  let _ = free x;
  load x // accessing x after free
\end{minted}

Contraction allows us to copy the x and use it as many times as we want. Similarly, nothing forces us to actually free the allocation thanks to Weakening. The above code type checks just fine.

However, if we change the type definitions to use linear types:
\begin{align*}
  &\vdash \texttt{alloc} : \tau \multimap \langle *\tau\rangle &
  &\vdash \texttt{load} : \langle *\tau\rangle \multimap \tau \\
  &\vdash \texttt{free} : \langle *\tau\rangle \multimap () &
  &\vdash \texttt{store} : \langle *\tau\rangle \multimap \tau \multimap ()
\end{align*}
The previous use after free code does \emph{not} type check any more, and a \(\langle *\tau\rangle\) \emph{uniquely} owns its memory. However, there is another problem, as these definitions are practically useless. The following code \emph{also} would not type check, because \mintinline{text}{x} is \emph{still} used twice:
\begin{minted}{text}
  let x = alloc v;
  let w = load x; // x is consumed here
  free x // does not exist anymore
\end{minted}

\subsubsection{Capability-Passing Style} \label{sec:theory:capability-passing-style}

To fix this issue, the usual solution is using different definitions of our functions utilizing something called \emph{capability-passing style}. In this style, functions that do something with a linear value without consuming them need to return the value back to the caller. That slightly changes the definitions of our memory management functions once again:
\begin{align*}
  &\vdash \texttt{alloc} : \tau \multimap \langle *\tau\rangle &
  &\vdash \texttt{load} : \langle *\tau\rangle \multimap \langle *\tau\rangle \otimes \tau \\
  &\vdash \texttt{free} : \langle *\tau\rangle \multimap () &
  &\vdash \texttt{store} : \langle *\tau\rangle \multimap \tau \multimap \langle *\tau\rangle
\end{align*}
Now the code from before can be written like so:
\begin{minted}{text}
  let x0 = alloc v;
  let (x1, w) = load x0;
  free x1 // new x1 so fine
\end{minted}

The \mintinline{text}{free} function still fully consumes the resource and \(\langle *\tau\rangle\) still uniquely owns its memory, yet \mintinline{text}{load} and \mintinline{text}{store} can now be used.

There are two issues with this approach however. Firstly, this style of programming is not very ergonomic, and secondly, nothing prevents the provider of the function from returning completely new unrelated objects. We will discuss how to fix these issues in \cref{sec:rust:borrowing}.

It is worth noting again that this is only a simplified view of the topic. \citet{wadler-linear-taste} e.g. goes into further detail.

\section{The Real World of Rust} \label{sec:rust}

In this section, we will explore how Rust uses and extends the ideas of linear types. Rust relies on the fact that a variable of a linear type is the unique owner of the value's memory. That way, Rust can prevent various memory safety related issues with the help of its type checker, completely at compile time, keeping the cost in runtime performance low. We will make use of several features implemented in Rust's standard library~\cite{rust:std} for the examples.

Many programming languages implement the ideas behind linear types in some ways. And while there are also languages explicitly implementing linear types~\cite{austral-intro, par-docs}, Rust is probably the most well known application of the concept.

Although that isn't quite accurate. Rust does in fact \emph{not} completely prevent Weakening. In particular, Rust can create memory leaks, and it is explicitly supported with \mintinline{rust}{std::mem::forget} (although this did cause issues in the past~\cite{rust:poop}). Instead, it may be more accurate to categorize Rust's type system as \emph{affine}. Affine logic is linear logic plus Weakening, or in other words, it is intuitionistic logic without Contraction. But Rust still employs some features that are inspired by the idea of disallowing Weakening.

\subsection{Rust's Ownership Model} \label{sec:rust:ownership}

We mentioned in \cref{sec:linear:logic} that linear logic doesn't just outright remove intuitionistic assumptions, but instead just groups assumptions into intuitionistic or linear. This categorization of assumptions shows itself in Rust's \Copy{} trait.\footnote{In a simplified view, traits can be viewed as Rust's type classes.}

\subsubsection{Non-\Copy{} Types}

Types that don't implement \Copy{} act like our linear types. To improve ergonomics, Rust uses the notion of \emph{Ownership} and \emph{move-semantics} for values of these types. A variable of a non-\Copy{} type \emph{owns} the corresponding value, and a value of a non-\Copy{} type can only ever have a single owner. In the following, we will use the \mintinline{text}{std::vec::Vec} type as an example for an owned type. The following code will be rejected by Rust's compiler:
\begin{minted}[linenos, mathescape]{rust}
  let mut v = vec![]; |\codelinelabel{line:non-copy:v}|
  let mut w = v; |\codelinelabel{line:non-copy:w}|
  v.push(1); // error $\codelinelabel{line:non-copy:e}$
  w.push(2);
\end{minted}
\Cref*{line:non-copy:v} defines the variable \mintinline{text}{v} that is the unique owner of the underlying vector. \Cref*{line:non-copy:w} \emph{moves} the value to \mintinline{text}{w}. Rust knows that the ownership of the underlying vector has moved from \mintinline{text}{v} to \mintinline{text}{w}, hence \mintinline{text}{v} cannot be used after \cref*{line:non-copy:w}. \Cref*{line:non-copy:e} causes an error.

Note that this code does not need Weakening to type. Similar to C++, Rust implicitly calls a destructor at the end of a variable's scope. In Rust terminology, this destructor function is called \mintinline{text}{drop}, and \mintinline{text}{std::vec::Vec}'s \mintinline{text}{drop} implementation would free whatever heap memory was allocated for the vector.

\subsubsection{\Copy{} Types}

Types implementing the \Copy{} marker trait represent values stored on the stack, which can be copied cheaply~\cite[Chapter 4.1]{rust:book}. While the type system doesn't actually prevent us from implementing \Copy{} on any type we want, it does prevent us from implementing both \Copy{} \emph{and} \mintinline{text}{Drop} on the same type~\cite{rust:ops-drop}.

If we replace the vector with an integer in the above code, the Rust compiler will accept it:
\begin{minted}[linenos]{rust}
  let mut v = 1;
  let mut w = v; // copies v
  v += 1; // ok because v is Copy
  w += 2;
\end{minted}

\subsection{Capability-Passing Style in Rust} \label{sec:rust:capability-passing-style}

\begin{figure*}
  \begin{subfigure}{0.46\textwidth}
    \begin{minted}[linenos]{rust}
      fn get_len(s: Vec<usize>) -> usize {
          s.len()
      }
      let v = vec![];
      let len = get_len(v);
      v.push(1); // error, v was moved
    \end{minted}
    \caption{Naive Implementation} \label{lst:get_len:naive}
  \end{subfigure}
  \begin{subfigure}{0.52\textwidth}
    \begin{minted}[linenos]{rust}
      fn get_len(s: Vec<usize>) -> (Vec<usize>, usize) {
          let len = s.len(); (s, len)
      }
      let v0 = vec![];
      let (mut v1, len) = get_len(v0);
      v1.push(1); // fine
    \end{minted}
    \caption{Capability-Passing Style} \label{lst:get_len:capability-passing-style}
  \end{subfigure}

  \vspace{1em}

  \begin{subfigure}{0.5\textwidth}
    \begin{minted}[linenos]{rust}
      fn gen_len(s: &Vec<usize>) -> usize { s.len() }
      let mut v = vec![];
      let len = gen_len(&v); // borrows v
      v.push(1);
    \end{minted}
    \caption{Using Borrows} \label{lst:get_len:borrows}
  \end{subfigure}
  \caption{\mintinline{text}{get_len} Example} \label{lst:get_len}
\end{figure*}

\Cref{lst:get_len:naive} shows how naively applying the rules of ownership results in a similar problem described in \cref{sec:theory:capability-passing-style}. Functions whose arguments are non-\Copy{} take ownership of those values. If one wants to use a value \emph{after} passing it to a function, the function must return the value, as seen in \cref{lst:get_len:capability-passing-style}.

As mentioned in \cref{sec:theory:capability-passing-style}, not only is this not ergonomic, the signature of this function is not very expressive. From \mintinline{text}{get_len}'s type signature alone, it is not clear that the returned vector is the same vector as the input. Nothing prevents \mintinline{text}{get_len} from dropping the provided vector and returning a newly allocated one. The following section describes Rust's approach to this problem.

\subsection{Borrowing} \label{sec:rust:borrowing}

The idea of borrowing builds upon the notion of references from languages like C++. As the name suggests, \emph{borrowing} a value to another piece of code (like a function) allows that other code limited access to the value, without the lender losing ownership.

There are two kinds of borrows: immutable and mutable borrows. Immutable borrows, indicated by a single \mintinline{text}{&}, allow read-only access to the underlying data, whereas mutable borrows, indicated by \mintinline{text}{&mut}, also allow mutation. The same \mintinline{text}{gen_len} function from before can now be expressed using an immutable borrow, as seen in \cref{lst:get_len:borrows}.

The usefulness of this approach goes beyond better ergonomics: A function argument that takes an immutable borrow also conveys a programmer's \emph{intent}: This function \emph{does not change the underlying object}, which is enforced by the compiler as well.
% TODO maybe mention parallel to const references?

Mutable borrows allow more operations, but are also governed by stricter rules than immutable borrows. The main difference is that an object can have any number of immutable borrows, but only ever a single mutable borrow. In particular, an object cannot have immutable and mutable borrows at the same time. Hence, immutable borrows are often referred to as "shared references", while mutable borrows are often called "exclusive references".

Viewing immutable borrows from the lens of the \mintinline{text}{alloc} example from \cref{sec:linear:types}, an immutable borrow of \(\langle *\tau\rangle\) is a type that sacrifices \mintinline{text}{free} and \mintinline{text}{store} in favour of reintroducing the Contraction and Weakening rules~\cite{from-linearity-to-borrowing}. In fact, Rust has a blanket implementation of the \Copy{} trait for all immutable references~\cite{rust:marker-copy} for this very purpose. On the other hand, a mutable borrow only sacrifices \mintinline{text}{free} in favour of Weakening.

\subsubsection{Exclusivity of Mutable Borrows}

The exclusivity of mutable borrows has three major motivations:
\begin{enumerate}
  \item Prevention of Aliasing \label{motivations:aliasing}
  \item Prevention of Data Races \label{motivations:data-races}
  \item Prevention of Dangling References \label{motivations:dangling}
\end{enumerate}

To illustrate \cref{motivations:aliasing}, we will use an example by \citet{rust:nomicon}, seen in \cref{lst:aliasing}. Rust implements references as pointers~\cite{rust:core-ref}, so the code in \cref{lst:aliasing:before-transformation}, naively viewed, is inefficient in that it has to dereference \mintinline{text}{input} twice. Ideally, the compiler would like to transform this code into the one shown in \cref{lst:aliasing:after-transformation} by first dereferencing the pointer, \enquote{caching} the underlying value and using that for the comparisons. However, this is \emph{only} possible if the two parameters do not alias. That is, the memory they point to cannot overlap. If they did, the block of the first \mintinline{text}{if} conditional may overwrite the value behind the \mintinline{text}{input} reference. By disallowing the creation of mutable and immutable references to the same object, and hence the same memory, at the same time, Rust also prevents aliasing of mutable and immutable references. So the optimization shown is sound in Rust, which mimics the behaviour of C's \mintinline{text}{restrict} keyword.

\begin{figure*}
  \centering
  \begin{subfigure}{0.42\textwidth}
    \centering
    \begin{minted}[linenos]{rust}
      fn compute(input: &u32,
                 output: &mut u32) {

          if *input > 10 { *output = 1; }
          if *input > 5 { *output *= 2; }
      }
    \end{minted}
    \caption{Before Transformation} \label{lst:aliasing:before-transformation}
  \end{subfigure}
  \begin{subfigure}{0.42\textwidth}
    \centering
    \begin{minted}[linenos]{rust}
      fn compute(input: &u32,
                 output: &mut u32) {
          let tmp = *input;
          if tmp > 10 { *output = 1; }
          if tmp > 5 { *output *= 2; }
      }
    \end{minted}
    \caption{After Transformation} \label{lst:aliasing:after-transformation}
  \end{subfigure}
  \caption{Aliasing Transformation Example} \label{lst:aliasing}
\end{figure*}

\Cref{motivations:data-races} is primarily a concern in concurrent applications. If we cannot create two mutable views of an object, we can't have two threads both writing to it at the same time either.

\Cref{motivations:dangling} is a more serious memory safety concern. See for example the following code, which Rust correctly refuses to compile:
\begin{minted}[linenos, mathescape]{rust}
  let mut v = vec![1, 2]; |\codelinelabel{line:dangling:v}|
  let m = &mut v;
  let x = &v[1]; // x points to value `2`
  m.clear(); // empties the vector behind m $\codelinelabel{line:dangling:clear}$
  println!("{x}");
\end{minted}
\Cref*{line:dangling:v} creates a vector with 2 elements called \mintinline{text}{v}. \mintinline{text}{m} is a mutable reference to \mintinline{text}{v}, while \mintinline{text}{x} is an immutable reference to \mintinline{text}{v}'s second element. Because \mintinline{text}{m} is a mutable reference, we can use it to clear the vector in \cref*{line:dangling:clear}, leaving \mintinline{text}{x} as a dangling reference to potentially freed memory.

The relevant issue is that, even though a mutable reference doesn't allow \emph{dropping} a value (mutable references still allow Weakening), mutation can result in memory being freed or moved. If other references were to alias that same memory, they may become dangling references.

\subsection{Lifetimes} \label{sec:rust:lifetimes}

So far, we could lend values to functions by creating borrows, but the space of possible return types was still limited to owned values. This is not a concern if one returns \Copy{} types, but becomes a concern once one tries to return some substructure of a non-\Copy{} type. For example, a typical operation on vectors is a \mintinline{text}{get} function, which takes an index and returns the value at that given index:
\begin{minted}[linenos]{rust}
  fn get(v: &Vec<usize>, index: usize)
      -> usize {
    v[index]
  }
\end{minted}

But \mintinline{text}{Vec} is a \emph{generic} data type. We can create vectors of any type \mintinline{text}{T} of which the size is known at compile time, so a function like \mintinline{text}{get} should be generic over the type of the vector:
\begin{minted}[linenos]{rust}
  fn get<T>(v: &Vec<T>, index: usize) -> T {
    v[index]
  }
\end{minted}
And herein lies the issue: \mintinline{text}{T} might not be \Copy{}. The return value \mintinline{rust}{-> T} implies that the function returns an \emph{owned} value of \mintinline{text}{T}. This, however, is not the case: the owner of the value we are trying to get to is the owner of the vector, which lies somewhere in the caller. We cannot take ownership of \mintinline{text}{v[index]}, because we do not have ownership of \mintinline{text}{v}. Of course, we could limit \mintinline{text}{get} to be only instantiatable for types \mintinline{text}{T} that \emph{are} \Copy{}, but this limits the usefulness of \mintinline{text}{get} significantly.

Instead, what one would like to do is be able to return borrows from functions like so:
\begin{minted}[linenos]{rust}
  fn get<T>(v: &Vec<T>, index: usize) -> &T {
    &v[index]
  }
\end{minted}
This code will actually compile with the Rust compiler today.

The issue with naively allowing references in function return types is that references don't have an infinite lifetime. A borrow must not live longer than the value's owner, or else there is a risk of a dangling reference which is alive even after the owner was dropped. This is where Rust's notion of \emph{lifetimes} come into play.

Every value has a lifetime. In its simplest form, a value's lifetime begins on the line where its constructor is called and ends when its destructor is called (i.e. when it goes out of scope and \mintinline{text}{drop} gets called). References also have lifetimes, which start where the borrow happens, and end on the reference's last use.~\cite{rust:nll-intro}

The fundamental rule is that a reference's lifetime must be entirely contained within the corresponding value's lifetime. This is the job of Rust's \emph{borrow checker}, a part of the compiler that makes sure this rule is never broken.

\subsubsection{Lifetime Parameters}

Lifetimes also come in the form of Lifetime parameters, and their names are of the form \enquote{\mintinline{text}{'a}}. These are similar to generic type parameters, only instead of quantifying over types, they quantify over lifetimes.

% TODO not sure how and if this is relevant, requires some better source probably (Oxide/RustBelt)
%
% One detail we skipped over in \cref{sec:theory} is that a generic function like \mintinline{rust}{fn alloc<T>(t: T) -> *T} can be seen as a universally quantified proposition \(\forall \texttt{T}\ldotp \texttt{T} \rightarrow *\texttt{T}\). In the same way, functions that are generic over lifetimes have a similar meaning: We are quantifying over all possible lifetimes.

The \mintinline{text}{get} function definition from \cref{sec:rust:lifetimes} made use of Rust's lifetime parameter elision rules, which are designed to reduce the number of cases where explicit lifetime annotations are required. We can make the lifetime parameters explicit:
\begin{minted}[linenos]{rust}
  fn get<'a, T>(v: &'a Vec<T>, index: usize)
      -> &'a T {
    &v[index]
  }
\end{minted}
\mintinline{text}{get} is not just generic over \mintinline{text}{T}, but also over the lifetime \mintinline{text}{'a}. When instantiating the function, both the type and the lifetime parameter must be instantiated to concrete types and lifetimes respectively. This new de-sugared function definition clearly states the \enquote{condition} that the returned borrow must fulfil: it's a reference that only lives as long as the borrow \mintinline{text}{&v}. The returned reference from \mintinline{text}{get} cannot be used \emph{after} the argument's lifetime ends.

This connection becomes clearer when looking at a caller of \mintinline{text}{get}:
\begin{minted}[linenos]{rust}
  let mut v = vec![1, 2];
  let x: &usize = get(&v, 1); |\tikzmark{lifetime-begin}\codelinelabel{line:lifetime:x}|
  v.push(3); |\codelinelabel{line:lifetime:push1}|
  v.push(4); |\codelinelabel{line:lifetime:push2}|
  println!("{x}"); |\tikzmark{lifetime-end}\codelinelabel{line:lifetime:xuse}|
\end{minted}
\begin{tikzpicture}[remember picture]
  \coordinate (begin) at (pic cs:lifetime-begin);
  \coordinate (begin-mid) at ($ (begin) + (0.5, 0) $);
  \coordinate (end) at (pic cs:lifetime-end);
  \coordinate (end-mid) at (begin-mid |- end);

  \draw[overlay, Stealth-Stealth, thick] (begin) -- (begin-mid) -- node[right] {lifetime \mintinline{text}{'a}} (end-mid) -- (end);
\end{tikzpicture}
This code will not compile. The borrow checker notices that \mintinline{text}{x} gets used in \cref*{line:lifetime:xuse}, thus the lifetime \mintinline{text}{'a} must include \emph{at least} \cref*{line:lifetime:x,line:lifetime:push1,line:lifetime:push2,line:lifetime:xuse}. The \mintinline{text}{get} function gets instantiated with this \mintinline{text}{'a} lifetime, and hence the lifetime of the \mintinline{text}{&v} borrow must also live at least until \cref*{line:lifetime:xuse}. Thus, an immutable borrow exists during \cref*{line:lifetime:push1,line:lifetime:push2}, so the mutation is not sound.

\subsubsection{Non-Lexical Lifetimes} \label{sec:rust:nll}

Prior to the 2018 edition of Rust (which was stabilized with Rust version 1.31~\cite{rust:2018-release}), the following code was not accepted by the borrow checker:
\begin{minted}[linenos]{rust}
  fn main() {
    let mut x = 5;
    let y: &usize = &x; |\codelinelabel{line:nll-mot:y}|
    x += 1; |\codelinelabel{line:nll-mot:mut}|
  }
\end{minted}
Even though we mutate \mintinline{text}{x} while an immutable borrow (\mintinline{text}{y}) exists to the underlying value, this code is completely safe, because \mintinline{text}{y} is never used after the mutation. Yet, the code used to give an error due to this aliasing. This is because lifetimes followed lexical scopes. \Cref*{line:nll-mot:y} creates an immutable borrow of \mintinline{text}{x}, its lifetime being bound to the scope of variable \mintinline{text}{y}. Because \mintinline{text}{y} stays in scope until the end of \mintinline{text}{main}, so does the immutable borrow.

Static analysis is inherently conservative, and the Rust compiler will always have cases where it rejects safe programs because of technical limitations of its analysis. The issue presented here, however, was solved with Rust edition 2018 with the stabilization of \emph{non-lexical lifetimes} (NLL for short)~\cite{rust:nll-intro, rust:nll2}. Instead of calculating lifetimes based on lexical scopes, the borrow checker today uses \emph{liveness\footnote{A variable is considered live at a point in the program, if it might be used sometime after that point.} analysis} of variables to define more concise lifetimes.

In a simplified view, a non-lexical lifetime is a set of lines\footnote{Or more accurately: a set of points in the control-flow graph.} where the lifetime is active, which is usually inferred by the set of lines where the corresponding variable is live. In the previous example, today's borrow checker will realize that \mintinline{text}{y} is not live after \cref*{line:nll-mot:y}, because it is never used. Hence, the lifetime of the borrow is completely confined to \cref*{line:nll-mot:y}. One could imagine the lifetime of that borrow is the set \(\{3\}\), representing only \cref*{line:nll-mot:y}. Not only that, but the \emph{type} of \mintinline{text}{y} will include that lifetime, as \mintinline{text}{y}'s type \emph{is} a borrow:
\mint{rust}{y: &{3} usize}
The mutation in \cref*{line:nll-mot:mut} thus doesn't overlap with the lifetime of an immutable borrow any more.

% Another example that wouldn't compile previously is a function that takes a mutable borrow of a \mintinline{text}{HashMap}, checks if some key is in the map, does some processing on the underlying value if it is, and inserts a default value if it isn't~\cite{rust:nll-intro}:
% \begin{minted}[linenos]{rust}
%   fn process_or_default(
%     map: &'a mut HashMap<usize, usize>,
%   ) {
%     match map.get_mut(&69) {
%       Some(v) => {
%         println!("{v}");
%         // ... processing v
%       }
%       None => {
%         map.insert(69, 420);
%         // ...
%       }
%     }
%   }
% \end{minted}

\subsection{Polonius} \label{sec:rust:polonius}

Even with NLL, there are still issues with today's borrow checker in practice. The typical example by \citet{rust:polonius-talk, rust:nll-intro, rust:nll2} is the attempt at a function that takes a mutable reference to a \mintinline{text}{HashMap}, returns the value behind some key, and inserts a default value for that key if the key isn't already present in the map (as in \cref{lst:get_or_insert:definition}).\footnote{Note that this specific example was a primary motivation for the development of Rust's \mintinline{text}{HashMap::entry} API~\cite{rust:std} which should be used for this specific use case instead.}

\begin{figure*}
  \begin{subfigure}[t]{0.48\textwidth}
    \begin{minted}[linenos]{rust}
      fn get_or_insert<'a>(
          map: &'a mut HashMap<usize, usize>,
          key: usize,
      ) -> &'a usize {
          match map.get(&key) { |\codelinelabel{line:get_or_insert:loan}|
              Some(v) => v, |\codelinelabel{line:get_or_insert:v}|
              None => {
                  map.insert(key, 420); |\codelinelabel{line:get_or_insert:mutation}|
                  &map[&key]
              }
          }
      }
    \end{minted}
    \caption{Definition} \label{lst:get_or_insert:definition}
  \end{subfigure}
  \begin{subfigure}[t]{0.4\textwidth}
    \begin{minted}[linenos, firstnumber=last]{rust}
      fn main() {
          let mut map = HashMap::new();
          let v = get_or_insert(&mut map, 69); |\codelinelabel{line:caller:call}|
          println!("{v}"); |\codelinelabel{line:caller:use}|
      }
    \end{minted}
    \caption{Caller} \label{lst:get_or_insert:caller}
  \end{subfigure}
  \caption{\mintinline{text}{get_or_insert} Lifetime Example} \label{lst:get_or_insert}
\end{figure*}

This function fails to compile on today's Rust, even with NLL. \mintinline{rust}{map.get(&key)} is short for \mintinline{rust}{HashMap::get(&*map, &key)}. In particular, \cref*{line:get_or_insert:loan} creates an immutable borrow of \mintinline{text}{map}. This borrow has a lifetime, and with the view of NLL, that lifetime corresponds to a set of lines where the borrow may still be used. However, the value returned by the call to \mintinline{text}{get} might actually be returned to the caller, so the lifetime of this borrow is not confined to this function. The lifetime applied to the borrow of map is thus \mintinline{text}{'a}.

\Cref{lst:get_or_insert:caller} shows a potential caller to the \mintinline{text}{get_or_insert} function. The lifetime that \mintinline{text}{get_or_insert} is instantiated with will be \(\{15, 16\}\), representing \cref*{line:caller:call,line:caller:use}, and \cref*{line:caller:call} actually includes the \emph{entire} call of \mintinline{text}{get_or_insert}. In \cref{sec:rust:nll} we mentioned that a non-lexical lifetime can be seen as a set of lines. However, it can also be a named lifetime like \mintinline{text}{'a}. In this example, the named lifetime \mintinline{text}{'a} from \mintinline{text}{get_or_insert} is a lifetime that exists somewhere in the caller, because the function \emph{returns a value of lifetime \mintinline{text}{'a}}. The lifetime gets blown up to the entirety of the \mintinline{text}{get_or_insert} function, and will include the \mintinline{text}{None} branch. When we now try to mutate inside the \mintinline{text}{None} branch (by calling \mintinline{text}{insert}), the borrow checker sees overlapping mutable and immutable borrows and errors out.

Polonius~\cite{rust:polonius-repo} is a new approach to lifetimes that tries to resolve this issue. With Polonius, a lifetime is not a set of lines where the borrow is live, but instead a set of \emph{loans} that the borrow might have come from, where a loan is the creation of a borrow. Analysing the simple example from \cref{sec:rust:nll}:
\begin{minted}[linenos]{rust}
  fn main() {
    let mut x = 5;
    let y: &usize = &x; |\codelinelabel{line:nll-ex:y}|
    x += 1; |\codelinelabel{line:nll-ex:mut}|
  }
\end{minted}
The lifetime of the borrow created at \cref*{line:nll-ex:y} is now the set of loans that the borrow may have come from. There is only one loan that fits and that is the loan done at \cref*{line:nll-ex:y}. The lifetime of \mintinline{text}{y} is thus \(\{\texttt{L}3\}\) and \mintinline{text}{y}'s type is \mintinline{rust}{y: &{L3} usize}. We now say a loan \(L\) is \emph{live}, if \emph{some live variable} has \(L\) in its type~\cite{rust:polonius-talk}.

Since \mintinline{text}{y} is never live, the loan from \cref*{line:nll-ex:y} isn't either, hence the mutation on \cref*{line:nll-ex:mut} is valid. If we were to add another line that is a use of \mintinline{text}{y}, \mintinline{text}{y} would become live, so would \(\texttt{L}3\), and the mutation on \cref*{line:nll-ex:mut} becomes invalid:
\begin{minted}[linenos, mathescape]{rust}
  fn main() {
    let mut x = 5;
    let y: &usize = &x;
    x += 1; // error, because y is live $\codelinelabel{line:nll-ex2:mut}$
    println!("{y}");
  }
\end{minted}
Previously with NLL, the lifetime (and hence type) of \mintinline{text}{y} would've changed. With Polonius, the lifetime and type of \mintinline{text}{y} stays unchanged. However, the liveness of \mintinline{text}{y} is still larger than before, and hence also the liveness of the \(\texttt{L}3\) loan. Since \(\texttt{L}3\) is live at \cref*{line:nll-ex2:mut}, that mutation violates the rules of the \(\texttt{L}3\) loan, and the program is rejected by the borrow checker.

Note the important distinction from NLL: With NLL, the liveness of the variable is encoded within its lifetime. With Polonius, the liveness of a variable is still required for the borrow checker, but the lifetimes themselves are decoupled entirely from liveness.

The \mintinline{text}{get_on_insert} example from \cref{lst:get_or_insert:definition} now also poses no problem, and is the main motivation behind Polonius's development. The original issue was that the loan of \mintinline{text}{map} in \cref*{line:get_or_insert:loan} got a lifetime that had to be set equal to \mintinline{text}{'a}, as that is the lifetime of the return value. \mintinline{text}{'a} is a named lifetime that spans the whole function. The lifetime of the loan from \cref*{line:get_or_insert:loan} thus overlapped with the mutation from \cref*{line:get_or_insert:mutation}.

With Polonius, we still have a loan at \cref*{line:get_or_insert:loan}, which we will call \(\texttt{L}5\). \(\texttt{L}5\) is part of \mintinline{text}{v}'s type, however, v is \emph{not live} at \cref*{line:get_or_insert:mutation}. Hence, the loan \(\texttt{L}5\) is \emph{also} not live at \cref*{line:get_or_insert:mutation}. Remember, with Polonius, the borrow checker needs to check if some instruction violates the rules of a loan that is live at that instruction. Since \mintinline{text}{v} is not live at \cref*{line:get_or_insert:mutation}, neither is \(\texttt{L}5\), and so there is no loan that is violated. Thus, the program gets accepted.

It is important to note that, at the time of writing, Polonius is a current development and not integrated in today's stable Rust compiler. The authors of Polonius themselves state, that it is not ready for widespread use~\cite{rust:polonius-book}.

% TODO probably got no space for this left
% \subsection{Lifetime Parameters for Structures}

\section{Summary}

Rust's type system has long evolved beyond its theoretical roots due to practical considerations, however its connection to linear types remains evident. In this paper, we displayed a limited view of Rust's ownership and borrowing rules, but many features and parts of Rust remain outside its scope. For example, contrary to the description from \cref{sec:rust:borrowing}, Rust \emph{does} support mutation through immutable borrows through the notion of interior mutability, which implements the borrow checking rules at runtime. Similarly, Rust's \mintinline{text}{unsafe} keyword allows certain operations, which permit circumventing the borrow checker for when its rules are too restrictive.

Several challenges with the borrow checker still remain. Some of them will improve by the eventual introduction of Polonius into the stable compiler, others remain unsolved. For instance, self-referential structs remain a pain point for many Rust developers and still require the use of awkward or unsafe APIs.

Ultimately however, the balance of safety guarantees and developer ergonomics offered by Rust is impressive. Even simple examples can cause complicated and intricate lifetime constraints, highlighting the extensive research and engineering effort required. By utilizing linear types, the language manages to offer the performance of systems programming languages with manual memory management, all the while rarely requiring manual intervention, as the type system figures out most on its own. All this is done in the type checker at compile time, causing virtually no runtime performance overhead.

% TODO maybe mention Polonius is written in Datalog?

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper} % read paper.bib file

\end{document}
